Draft Final Report: Measles risk assessment, modelling, and benefit--cost analyses for New Zealand

### 2014-11-07 draftfinalreport.tex
edited final draft

This large number of categories, some with small population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes. Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of \textit{individual} NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from ten to two: NZDep 1--5 and NZDep 6--10. We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category. By doing so, we reduce the zeroinflation present in the data.

The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories. For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:case}).

\begin{table}[hbtp]
\footnotesize
%latex.default(ttyr, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllr}
\hline\hline
\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Cases}\tabularnewline
\hline
1-5&0-2&Asian&$ 11$\tabularnewline
6-10&0-2&Asian&$  8$\tabularnewline
1-5&3-5&Asian&$  1$\tabularnewline
1-5&6-17&Asian&$ 11$\tabularnewline
6-10&6-17&Asian&$  5$\tabularnewline
1-5&18-24&Asian&$  3$\tabularnewline
6-10&18-24&Asian&$  5$\tabularnewline
1-5&25+&Asian&$ 10$\tabularnewline
6-10&25+&Asian&$ 13$\tabularnewline
1-5&0-2&European&$ 83$\tabularnewline
6-10&0-2&European&$ 64$\tabularnewline
1-5&3-5&European&$ 42$\tabularnewline
6-10&3-5&European&$ 17$\tabularnewline
1-5&6-17&European&$219$\tabularnewline
6-10&6-17&European&$ 80$\tabularnewline
1-5&18-24&European&$ 34$\tabularnewline
6-10&18-24&European&$ 36$\tabularnewline
1-5&25+&European&$ 78$\tabularnewline
6-10&25+&European&$ 51$\tabularnewline
1-5&0-2&Maori&$ 18$\tabularnewline
6-10&0-2&Maori&$ 48$\tabularnewline
1-5&3-5&Maori&$  7$\tabularnewline
6-10&3-5&Maori&$ 11$\tabularnewline
1-5&6-17&Maori&$ 19$\tabularnewline
6-10&6-17&Maori&$ 92$\tabularnewline
1-5&18-24&Maori&$  5$\tabularnewline
6-10&18-24&Maori&$  8$\tabularnewline
1-5&25+&Maori&$  2$\tabularnewline
6-10&25+&Maori&$  6$\tabularnewline
6-10&0-2&MLA&$  3$\tabularnewline
6-10&3-5&MLA&$  1$\tabularnewline
6-10&6-17&MLA&$  1$\tabularnewline
6-10&18-24&MLA&$  2$\tabularnewline
6-10&25+&MLA&$  6$\tabularnewline
1-5&0-2&Pacific&$  5$\tabularnewline
6-10&0-2&Pacific&$ 58$\tabularnewline
6-10&3-5&Pacific&$  3$\tabularnewline
1-5&6-17&Pacific&$  5$\tabularnewline
6-10&6-17&Pacific&$ 22$\tabularnewline
1-5&18-24&Pacific&$  1$\tabularnewline
6-10&18-24&Pacific&$  8$\tabularnewline
1-5&25+&Pacific&$  2$\tabularnewline
6-10&25+&Pacific&$ 11$\tabularnewline
1-5&0-2&None&$  3$\tabularnewline
1-5&3-5&None&$  1$\tabularnewline
1-5&6-17&None&$  3$\tabularnewline
6-10&6-17&None&$  4$\tabularnewline
1-5&18-24&None&$  2$\tabularnewline
6-10&18-24&None&$  1$\tabularnewline
1-5&25+&None&$  5$\tabularnewline
6-10&25+&None&$  3$\tabularnewline
\hline
\end{tabular}\end{center}\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
\label{table:case}
\end{table}

The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}. Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses. However, there are numerous issues with the data for Middle Eastern, Latin American and African ethnicities category, and along with small population sizes (Table~\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table~\ref{table:percap}).

\begin{figure}
\begin{center}
\includegraphics{draftfinalreport-010}
\end{center}
\caption{Per capita rates of measles infections for 2007--2014 broken down by ethnicity (Table~\ref{table:percap} for details)}
\label{fig:percap}
\end{figure}

\begin{table}[hbtp]
\footnotesize
\caption{Numbers of measles cases, population sizes and per capita rates of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
%latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllrrr}
\hline\hline
\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Population}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per capita}\tabularnewline
\hline
1-5&0-2&Asian&$   6094$&$ 11$&$0.0018$\tabularnewline
6-10&0-2&Asian&$   6806$&$  8$&$0.0012$\tabularnewline
1-5&3-5&Asian&$   6094$&$  1$&$0.0002$\tabularnewline
6-10&3-5&Asian&$   6806$&$  0$&$0.0000$\tabularnewline
1-5&6-17&Asian&$  33918$&$ 11$&$0.0003$\tabularnewline
6-10&6-17&Asian&$  28905$&$  5$&$0.0002$\tabularnewline
1-5&18-24&Asian&$  22917$&$  3$&$0.0001$\tabularnewline
6-10&18-24&Asian&$  34107$&$  5$&$0.0001$\tabularnewline
1-5&25+&Asian&$  96357$&$ 10$&$0.0001$\tabularnewline
6-10&25+&Asian&$  98715$&$ 13$&$0.0001$\tabularnewline
1-5&0-2&European&$  57872$&$ 83$&$0.0014$\tabularnewline
6-10&0-2&European&$  45445$&$ 64$&$0.0014$\tabularnewline
1-5&3-5&European&$  57872$&$ 42$&$0.0007$\tabularnewline
6-10&3-5&European&$  45445$&$ 17$&$0.0004$\tabularnewline
1-5&6-17&European&$ 264330$&$219$&$0.0008$\tabularnewline
6-10&6-17&European&$ 182937$&$ 80$&$0.0004$\tabularnewline
1-5&18-24&European&$ 107649$&$ 34$&$0.0003$\tabularnewline
6-10&18-24&European&$ 117840$&$ 36$&$0.0003$\tabularnewline
1-5&25+&European&$1001916$&$ 78$&$0.0001$\tabularnewline
6-10&25+&European&$ 724317$&$ 51$&$0.0001$\tabularnewline
1-5&0-2&Maori&$  10003$&$ 18$&$0.0018$\tabularnewline
6-10&0-2&Maori&$  30104$&$ 48$&$0.0016$\tabularnewline
1-5&3-5&Maori&$  10003$&$  7$&$0.0007$\tabularnewline
6-10&3-5&Maori&$  30104$&$ 11$&$0.0004$\tabularnewline
1-5&6-17&Maori&$  40461$&$ 19$&$0.0005$\tabularnewline
6-10&6-17&Maori&$ 116640$&$ 92$&$0.0008$\tabularnewline
1-5&18-24&Maori&$  15360$&$  5$&$0.0003$\tabularnewline
6-10&18-24&Maori&$  48495$&$  8$&$0.0002$\tabularnewline
1-5&25+&Maori&$  71217$&$  2$&$0.0000$\tabularnewline
6-10&25+&Maori&$ 192729$&$  6$&$0.0000$\tabularnewline
1-5&0-2&MLA&$    728$&$  0$&$0.0000$\tabularnewline
6-10&0-2&MLA&$   1290$&$  3$&$0.0023$\tabularnewline
1-5&3-5&MLA&$    728$&$  0$&$0.0000$\tabularnewline
6-10&3-5&MLA&$   1290$&$  1$&$0.0008$\tabularnewline
1-5&6-17&MLA&$   2991$&$  0$&$0.0000$\tabularnewline
6-10&6-17&MLA&$   4539$&$  1$&$0.0002$\tabularnewline
1-5&18-24&MLA&$   1710$&$  0$&$0.0000$\tabularnewline
6-10&18-24&MLA&$   3078$&$  2$&$0.0006$\tabularnewline
1-5&25+&MLA&$   8028$&$  0$&$0.0000$\tabularnewline
6-10&25+&MLA&$  10335$&$  6$&$0.0006$\tabularnewline
1-5&0-2&Pacific&$   2093$&$  5$&$0.0024$\tabularnewline
6-10&0-2&Pacific&$  13124$&$ 58$&$0.0044$\tabularnewline
1-5&3-5&Pacific&$   2093$&$  0$&$0.0000$\tabularnewline
6-10&3-5&Pacific&$  13124$&$  3$&$0.0002$\tabularnewline
1-5&6-17&Pacific&$   8541$&$  5$&$0.0006$\tabularnewline
6-10&6-17&Pacific&$  51183$&$ 22$&$0.0004$\tabularnewline
1-5&18-24&Pacific&$   3972$&$  1$&$0.0003$\tabularnewline
6-10&18-24&Pacific&$  22098$&$  8$&$0.0004$\tabularnewline
1-5&25+&Pacific&$  18492$&$  2$&$0.0001$\tabularnewline
6-10&25+&Pacific&$  91533$&$ 11$&$0.0001$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:percap}
\end{table}

@@

\begin{equation} \label{eq:reg}
 log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a{}_e(x_a * x_e)+ log(population)  + \epsilon
  \end{equation}
Where $\alpha$ is the intercept, $y$ cases, $_a$ age, $_e$ Prioritized Ethnicity, $_n$ NZDep, and $\epsilon$ the error term.

\subsection{Vaccination history of measles infection and population immunity estimation methods}

In this section we describe the vaccination history of the measles cases from 2007--2014 outbreaks using the data provided by ESR. We describe the population immunity levels at both a national level and that for each DHB in New Zealand. To do this we use the census data from NZ statistics (Statistics New Zealand, 2013) and the vaccination and serosurvey data provided by the Ministry of Health at the commencement of this work. Specifically, we use the data in Table \ref{table:sero}, that uses the serosurvey results and the National Immunisation Register for under 6 year olds.

@@

For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel data by country and year to measure human movement to and from New Zealand  (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}). We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose, and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports. We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from. We used simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation. We use the data from 2012 because this year had the most complete WHO measles data and yet was most recent, thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.

@@

The predicted values from the regression model plotted against the reported cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
\begin{figure}
\begin{center}
\includegraphics{draftfinalreport-015}
\end{center}
\caption{Regression model (Equation~\ref{eq:reg}) predictions plotted against the cases (Table~\ref{table:percap})}
\label{fig:predict}
\end{figure}

@@

The significance of the different predictor variables can be seen in the ANOVA results (Table~\ref{table:anova}).
\par
\vspace{5mm} %5mm vertical space
\begin{table}
% latex table generated in R 3.0.2 by xtable 1.7-1 package
% Mon Nov 03 20:46:33 2014
\begin{tabular}{lrrrrrr}
  \hline
 & Df & Deviance & Resid. Df & Resid. Dev & F & Pr($>$F) \\ 
  \hline
NULL &  &  & 39 & 1453.62 &  &  \\ 
  Age & 4 & 1304.20 & 35 & 149.43 & 183.58 & 0.0000 \\ 
  Ethnicity & 3 & 20.00 & 32 & 129.43 & 3.75 & 0.0285 \\ 
  NZDep & 1 & 10.70 & 31 & 118.74 & 6.02 & 0.0239 \\ 
  Age:Ethnicity & 12 & 81.91 & 19 & 36.83 & 3.84 & 0.0045 \\ 
   \hline
\end{tabular}\caption{Significance of different predictor variables for measles risk factors from 2007-2014}
\label{table:anova}
\end{table}
\vspace{5mm} %5mm vertical space
A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in Table~\ref{table:regression}.

\vspace{5mm} %5mm vertical space
% latex table generated in R 3.0.2 by xtable 1.7-1 package
% Mon Nov 03 20:46:33 2014
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & -6.4674 & 0.1148 & -56.32 & 0.0000 \\ 
  Age3-5 & -0.9129 & 0.2054 & -4.44 & 0.0003 \\ 
  Age6-17 & -0.7618 & 0.1343 & -5.67 & 0.0000 \\ 
  Age18-24 & -1.5050 & 0.1937 & -7.77 & 0.0000 \\ 
  Age25+ & -2.9507 & 0.1608 & -18.35 & 0.0000 \\ 
  EthnicityAsian & 0.0530 & 0.3250 & 0.16 & 0.8721 \\ 
  EthnicityMaori & 0.2124 & 0.1994 & 1.07 & 0.2999 \\ 
  EthnicityPacific & 1.1603 & 0.2043 & 5.68 & 0.0000 \\ 
  NZDep6-10 & -0.2119 & 0.0855 & -2.48 & 0.0227 \\ 
  Age3-5:EthnicityAsian & -2.0315 & 1.3827 & -1.47 & 0.1581 \\ 
  Age6-17:EthnicityAsian & -1.0074 & 0.4717 & -2.14 & 0.0459 \\ 
  Age18-24:EthnicityAsian & -0.8311 & 0.5941 & -1.40 & 0.1779 \\ 
  Age25+:EthnicityAsian & 0.4211 & 0.4433 & 0.95 & 0.3541 \\ 
  Age3-5:EthnicityMaori & -0.3864 & 0.4096 & -0.94 & 0.3573 \\ 
  Age6-17:EthnicityMaori & -0.0855 & 0.2469 & -0.35 & 0.7328 \\ 
  Age18-24:EthnicityMaori & -0.5828 & 0.4484 & -1.30 & 0.2092 \\ 
  Age25+:EthnicityMaori & -1.0482 & 0.5242 & -2.00 & 0.0600 \\ 
  Age3-5:EthnicityPacific & -2.1316 & 0.8139 & -2.62 & 0.0169 \\ 
  Age6-17:EthnicityPacific & -1.4541 & 0.3347 & -4.35 & 0.0003 \\ 
  Age18-24:EthnicityPacific & -0.9827 & 0.5129 & -1.92 & 0.0705 \\ 
  Age25+:EthnicityPacific & -0.6127 & 0.4367 & -1.40 & 0.1767 \\ 
   \hline
\end{tabular}\caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
\label{table:regression}
\end{table}
\vspace{5mm} %5mm vertical space

@@

People of Pacific origin are also over-represented as measles cases ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes. In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected. The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}). This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:}). Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.

@@

The regression analyses suggest that age is a particularly strong risk factor for measles. This comes as no surpise to epidemiologists or health care providers. However, our analyses also highlight other groups that are at greater risk of measles. In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age. Interpretation of these results must still be viewed with some caution, however, as there is very likely a spatial effect that might not be accounted for in these analyses. Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the National Immunisation Register (NIR). A key issue with incorporating the spatial immunisation data has been the denominator data and the NIR data. The data suggest that census data and NIR data re recording children living in different DHB to that which they are vaccinated in, or subsequent to the census or NIR data moving. The leads to some area census units having more than 100\% children vaccinated, with some many times more, and some with very low levels reportedly vaccinated. Another issue was how to deal with people of greater age than those recorded in the NIR. Thus, these analyses are possible future directions for this work and a focus of future data collection could be to better understand the link between denominator data and NIR data.

>>>>

This large number of categories, some with small population sizes, leads to both overdispersion and zeroinflation, as there are many categories with zero cases in, particularly in the adult age classes. Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of \textit{individual} NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from ten to two: NZDep 1--5 (least deprived) and NZDep 6--10 (most deprived) (\href{http://www.otago.ac.nz/wellington/research/hirp/otago020194.html}{www.otago.ac.nz}). We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category. By doing so, we reduce the zeroinflation present in the data.

The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories. For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:none}).

\begin{table}[hbtp]
\footnotesize
%latex.default(ttyr, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllr}
\hline\hline
\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Cases}\tabularnewline
\hline
1-5&0-2&None&$3$\tabularnewline
1-5&3-5&None&$1$\tabularnewline
1-5&6-17&None&$3$\tabularnewline
6-10&6-17&None&$4$\tabularnewline
1-5&18-24&None&$2$\tabularnewline
6-10&18-24&None&$1$\tabularnewline
1-5&25+&None&$5$\tabularnewline
6-10&25+&None&$3$\tabularnewline
\hline
\end{tabular}\end{center}\caption{Absolute number of measles cases in specific age and socio-economic deprivation categories from 2007-2014 with no ethnicity recorded}
\label{table:none}
\end{table}

The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}. Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses. However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see Table~\ref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table~\ref{table:percap}).

\begin{figure}
\begin{center}
\includegraphics{draftfinalreport-010}
\end{center}
\caption{Per capita rates of measles infections for specific age and New Zealand deprivation indices (NZDep) for 2007--2014, broken down by ethnicity (Table~\ref{table:percap} for details)}
\label{fig:percap}
\end{figure}

\begin{table}[hbtp]
\footnotesize
\caption{Numbers of measles cases, population sizes and per capita rates per 10,000 of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
%latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllrrr}
\hline\hline
\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Population}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per capita}\tabularnewline
\hline
1-5&0-2&Asian&$   6094$&$ 11$&$18.0505$\tabularnewline
6-10&0-2&Asian&$   6806$&$  8$&$11.7543$\tabularnewline
1-5&3-5&Asian&$   6094$&$  1$&$ 1.6410$\tabularnewline
6-10&3-5&Asian&$   6806$&$  0$&$ 0.0000$\tabularnewline
1-5&6-17&Asian&$  33918$&$ 11$&$ 3.2431$\tabularnewline
6-10&6-17&Asian&$  28905$&$  5$&$ 1.7298$\tabularnewline
1-5&18-24&Asian&$  22917$&$  3$&$ 1.3091$\tabularnewline
6-10&18-24&Asian&$  34107$&$  5$&$ 1.4660$\tabularnewline
1-5&25+&Asian&$  96357$&$ 10$&$ 1.0378$\tabularnewline
6-10&25+&Asian&$  98715$&$ 13$&$ 1.3169$\tabularnewline
1-5&0-2&European&$  57872$&$ 83$&$14.3420$\tabularnewline
6-10&0-2&European&$  45445$&$ 64$&$14.0830$\tabularnewline
1-5&3-5&European&$  57872$&$ 42$&$ 7.2574$\tabularnewline
6-10&3-5&European&$  45445$&$ 17$&$ 3.7408$\tabularnewline
1-5&6-17&European&$ 264330$&$219$&$ 8.2851$\tabularnewline
6-10&6-17&European&$ 182937$&$ 80$&$ 4.3731$\tabularnewline
1-5&18-24&European&$ 107649$&$ 34$&$ 3.1584$\tabularnewline
6-10&18-24&European&$ 117840$&$ 36$&$ 3.0550$\tabularnewline
1-5&25+&European&$1001916$&$ 78$&$ 0.7785$\tabularnewline
6-10&25+&European&$ 724317$&$ 51$&$ 0.7041$\tabularnewline
1-5&0-2&Maori&$  10003$&$ 18$&$17.9946$\tabularnewline
6-10&0-2&Maori&$  30104$&$ 48$&$15.9447$\tabularnewline
1-5&3-5&Maori&$  10003$&$  7$&$ 6.9979$\tabularnewline
6-10&3-5&Maori&$  30104$&$ 11$&$ 3.6540$\tabularnewline
1-5&6-17&Maori&$  40461$&$ 19$&$ 4.6959$\tabularnewline
6-10&6-17&Maori&$ 116640$&$ 92$&$ 7.8875$\tabularnewline
1-5&18-24&Maori&$  15360$&$  5$&$ 3.2552$\tabularnewline
6-10&18-24&Maori&$  48495$&$  8$&$ 1.6497$\tabularnewline
1-5&25+&Maori&$  71217$&$  2$&$ 0.2808$\tabularnewline
6-10&25+&Maori&$ 192729$&$  6$&$ 0.3113$\tabularnewline
1-5&0-2&MLA&$    728$&$  0$&$ 0.0000$\tabularnewline
6-10&0-2&MLA&$   1290$&$  3$&$23.2558$\tabularnewline
1-5&3-5&MLA&$    728$&$  0$&$ 0.0000$\tabularnewline
6-10&3-5&MLA&$   1290$&$  1$&$ 7.7519$\tabularnewline
1-5&6-17&MLA&$   2991$&$  0$&$ 0.0000$\tabularnewline
6-10&6-17&MLA&$   4539$&$  1$&$ 2.2031$\tabularnewline
1-5&18-24&MLA&$   1710$&$  0$&$ 0.0000$\tabularnewline
6-10&18-24&MLA&$   3078$&$  2$&$ 6.4977$\tabularnewline
1-5&25+&MLA&$   8028$&$  0$&$ 0.0000$\tabularnewline
6-10&25+&MLA&$  10335$&$  6$&$ 5.8055$\tabularnewline
1-5&0-2&Pacific&$   2093$&$  5$&$23.8892$\tabularnewline
6-10&0-2&Pacific&$  13124$&$ 58$&$44.1938$\tabularnewline
1-5&3-5&Pacific&$   2093$&$  0$&$ 0.0000$\tabularnewline
6-10&3-5&Pacific&$  13124$&$  3$&$ 2.2859$\tabularnewline
1-5&6-17&Pacific&$   8541$&$  5$&$ 5.8541$\tabularnewline
6-10&6-17&Pacific&$  51183$&$ 22$&$ 4.2983$\tabularnewline
1-5&18-24&Pacific&$   3972$&$  1$&$ 2.5176$\tabularnewline
6-10&18-24&Pacific&$  22098$&$  8$&$ 3.6202$\tabularnewline
1-5&25+&Pacific&$  18492$&$  2$&$ 1.0815$\tabularnewline
6-10&25+&Pacific&$  91533$&$ 11$&$ 1.2018$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:percap}
\end{table}

@@

\begin{equation} \label{eq:reg}
 \log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a{}_e(x_a \times x_e)+ \log(population)  + \epsilon
  \end{equation}
Where $\alpha$ is the intercept, $y$ signifies cases, $_a$ signifies age, $_e$ signifies Prioritized Ethnicity, $_n$ signifies NZDep, and $\epsilon$ is the error term.

\subsection{Vaccination history of measles infection and population immunity estimation methods}
\label{sub:immunestat}

In this section we describe the vaccination history of the measles cases from 2007--2014 outbreaks using the data provided by ESR. We describe the population immunity levels at both a national level and for each DHB in New Zealand. To do this we use the census data from NZ statistics (Statistics New Zealand, 2013) and the vaccination and serosurvey data provided by the Ministry of Health at the commencement of this work. Specifically, we use the data in Table \ref{table:sero}, derived from the serosurvey results and the National Immunisation Register (NIR) for under 6 year olds.

@@

For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel destination data by country and year, to measure human movement to and from New Zealand  (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}). We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose, and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports. We use the WHO data to determine per capita measles cases for each year and use these data and the number of immigrants to New Zealand to begin to understand where measles is likely to be imported from. We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation. We use the data from 2012 because this year had the most complete WHO measles data and yet is most recent, thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.

@@

The predicted values from the regression model plotted against the reported cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
\begin{figure}
\begin{center}
\includegraphics{draftfinalreport-015}
\end{center}
\caption{Regression model (Equation~\ref{eq:reg}) predictions plotted against the case data (Table~\ref{table:percap})}
\label{fig:predict}
\end{figure}

@@

The significance of the different predictor variables can be seen in the ANOVA results (Table~\ref{table:anova}).
\par
\vspace{5mm} %5mm vertical space
\begin{table}
% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Fri Nov 07 22:00:40 2014
\begin{tabular}{lrrrrrr}
  \hline
 & Df & Deviance & Resid. Df & Resid. Dev & F & Pr($>$F) \\ 
  \hline
NULL &  &  & 39 & 1453.62 &  &  \\ 
  Age & 4 & 1304.20 & 35 & 149.43 & 183.58 & 0.0000 \\ 
  Ethnicity & 3 & 20.00 & 32 & 129.43 & 3.75 & 0.0285 \\ 
  NZDep & 1 & 10.70 & 31 & 118.74 & 6.02 & 0.0239 \\ 
  Age:Ethnicity & 12 & 81.91 & 19 & 36.83 & 3.84 & 0.0045 \\ 
   \hline
\end{tabular}\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using Equation~\ref{eq:reg} and a quasipoisson error structure}
\label{table:anova}
\end{table}
\vspace{5mm} %5mm vertical space
A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in Table~\ref{table:regression}.

\vspace{5mm} %5mm vertical space
% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Fri Nov 07 22:00:40 2014
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & -6.4674 & 0.1148 & -56.32 & 0.0000 \\ 
  Age3-5 & -0.9129 & 0.2054 & -4.44 & 0.0003 \\ 
  Age6-17 & -0.7618 & 0.1343 & -5.67 & 0.0000 \\ 
  Age18-24 & -1.5050 & 0.1937 & -7.77 & 0.0000 \\ 
  Age25+ & -2.9507 & 0.1608 & -18.35 & 0.0000 \\ 
  EthnicityAsian & 0.0530 & 0.3250 & 0.16 & 0.8721 \\ 
  EthnicityMaori & 0.2124 & 0.1994 & 1.07 & 0.2999 \\ 
  EthnicityPacific & 1.1603 & 0.2043 & 5.68 & 0.0000 \\ 
  NZDep6-10 & -0.2119 & 0.0855 & -2.48 & 0.0227 \\ 
  Age3-5:EthnicityAsian & -2.0315 & 1.3827 & -1.47 & 0.1581 \\ 
  Age6-17:EthnicityAsian & -1.0074 & 0.4717 & -2.14 & 0.0459 \\ 
  Age18-24:EthnicityAsian & -0.8311 & 0.5941 & -1.40 & 0.1779 \\ 
  Age25+:EthnicityAsian & 0.4211 & 0.4433 & 0.95 & 0.3541 \\ 
  Age3-5:EthnicityMaori & -0.3864 & 0.4096 & -0.94 & 0.3573 \\ 
  Age6-17:EthnicityMaori & -0.0855 & 0.2469 & -0.35 & 0.7328 \\ 
  Age18-24:EthnicityMaori & -0.5828 & 0.4484 & -1.30 & 0.2092 \\ 
  Age25+:EthnicityMaori & -1.0482 & 0.5242 & -2.00 & 0.0600 \\ 
  Age3-5:EthnicityPacific & -2.1316 & 0.8139 & -2.62 & 0.0169 \\ 
  Age6-17:EthnicityPacific & -1.4541 & 0.3347 & -4.35 & 0.0003 \\ 
  Age18-24:EthnicityPacific & -0.9827 & 0.5129 & -1.92 & 0.0705 \\ 
  Age25+:EthnicityPacific & -0.6127 & 0.4367 & -1.40 & 0.1767 \\ 
   \hline
\end{tabular}\caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
\label{table:regression}
\end{table}
\vspace{5mm} %5mm vertical space

@@

The majority of measles cases are among people of European ancestry (Table~\ref{table:percap} and Figure~\ref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001). Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes. In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected. The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}). This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:percap}). Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.

@@

The regression analyses suggest that age is a particularly strong risk factor for measles. This comes as no surpise to epidemiologists or health care providers. However, our analyses also highlight other groups that are at greater risk of measles. In particular, Pacific people are at greater risk, as are the less deprived (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age. Interpretation of these results must still be viewed with some caution, however, as there is very likely a spatial effect that might not be accounted for in these analyses. Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the NIR. A key issue with incorporating the spatial immunisation data has been the denominator data and the NIR data. The data suggest that census data and NIR data are recording children living in different DHBs to that in which they are vaccinated, or subsequent to the census or NIR data moving. This leads to some area census units having more than 100\% children vaccinated, with some many times more, and some with very low levels reportedly vaccinated. Another issue was how to deal with people of greater age than those recorded in the NIR. Thus, these analyses are possible future directions for this work and a focus of future data collection could be to better understand the link between denominator data and NIR data.
